On the Limitations of Weighted Model Counting

### 2020-03-14 paper.tex
Comparison to other encodings

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
\begin{proposition}
  Let $\mathbf{B}$ and $\mathbf{B'}$ be Boolean algebras, and $\iota :
  \mathbf{B} \to \mathbf{B'}$ be the inclusion map such that $\mathbf{B}$ is
  freely generated by $L$, all atoms of $\mathbf{B'}$ can be expressed as
  meets of elements of $L'$, and the following subset relations are satisfied:
  \[
    \begin{tikzcd}
      \mathbf{B} \arrow{r}{\iota} & \mathbf{B'} \\
      L \arrow[Subset]{u}{} \arrow[Subset]{r}{} & L' \arrow[Subset]{u}{}
    \end{tikzcd}
  \]
  If, for any measure $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$, one can
  construct a weight function $w : L' \to \mathbb{R}_{>0}$ such that the WMC
  measure $\WMC : \mathbf{B'} \to \mathbb{R}_{\ge 0}$ with respect to $w$
  satisfies
  \[
    m = \WMC \circ \iota,
  \]
  then $|L' \setminus L| \ge 2^{|L|+1}$.
\end{proposition}
\begin{proof}
  % 1. An atom in B' must have more than just elements of L.
  Let $a$ be an atom in $\mathbf{B}$, and let $b$ be an atom in $\mathbf{B'}$
  such that $b \le a$. First, let us notice that as long as $|L| \ge
  4$\footnote{Note that $|L|$ has to be an even number.}, $b \ne a$. Indeed, let
  $p, r, \neg p, \neg r \in L$. Then
  \begin{align*}
    (\WMC \circ \iota)(p \land r) &= w(p)w(r), \\
    (\WMC \circ \iota)(p \land \neg r) &= w(p)w(\neg r), \\
    (\WMC \circ \iota)(\neg p \land r) &= w(\neg p)w(r), \\
    (\WMC \circ \iota)(\neg p \land \neg r) &= w(\neg p)w(\neg r), \\
  \end{align*}
  But then we have that
  \[
    \frac{m(p \land r)}{m(\neg p \land r)} = \frac{w(p)}{w(\neg p)} =
    \frac{m(p \land \neg r)}{m(\neg p \land \neg r)}.
  \]
  This places a condition on $m$, contradicting the assumption that the
  construction works for an arbitrary $m$. Hence $b < a$.

  % 2. If we add phi, then we must also add -phi.
  % 3. Extension to multiple literals: we must have all (2^n) combinations of
  % added literals).
  % 4. Profit
\end{proof}

% TODO: describe how my approach is better in the worst case than that by
% Darwiche
% TODO: what about ENC2?
Let us note how our lower bound on the number of added literals compares to ENC1
\cite{DBLP:conf/kr/Darwiche02} and ENC2

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

\begin{conjecture}
  Let $\mathbf{B}$ and $\mathbf{B'}$ be Boolean algebras, and $\iota :
  \mathbf{B} \to \mathbf{B'}$ be the inclusion map such that $\mathbf{B}$ is
  freely generated by $L$, all atoms of $\mathbf{B'}$ can be expressed as
  meets of elements of $L'$, and the following subset relations are satisfied:
  \[
    \begin{tikzcd}
      \mathbf{B} \arrow{r}{\iota} & \mathbf{B'} \\
      L \arrow[Subset]{u}{} \arrow[Subset]{r}{} & L' \arrow[Subset]{u}{}
    \end{tikzcd}
  \]
  If, for any measure $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$, one can
  construct a weight function $w : L' \to \mathbb{R}_{>0}$ such that the WMC
  measure $\WMC : \mathbf{B'} \to \mathbb{R}_{\ge 0}$ with respect to $w$
  satisfies
  \[
    m = \WMC \circ \iota,
  \]
  then $|L' \setminus L| \ge 2^{|L|+1}$.
\end{conjecture}
% \begin{proof}
%   % 1. An atom in B' must have more than just elements of L.
%   Let $a$ be an atom in $\mathbf{B}$, and let $b$ be an atom in $\mathbf{B'}$
%   such that $b \le a$. First, let us notice that as long as $|L| \ge
%   4$\footnote{Note that $|L|$ has to be an even number.}, $b \ne a$. Indeed, let
%   $p, r, \neg p, \neg r \in L$. Then
%   \begin{align*}
%     (\WMC \circ \iota)(p \land r) &= w(p)w(r), \\
%     (\WMC \circ \iota)(p \land \neg r) &= w(p)w(\neg r), \\
%     (\WMC \circ \iota)(\neg p \land r) &= w(\neg p)w(r), \\
%     (\WMC \circ \iota)(\neg p \land \neg r) &= w(\neg p)w(\neg r), \\
%   \end{align*}
%   But then we have that
%   \[
%     \frac{m(p \land r)}{m(\neg p \land r)} = \frac{w(p)}{w(\neg p)} =
%     \frac{m(p \land \neg r)}{m(\neg p \land \neg r)}.
%   \]
%   This places a condition on $m$, contradicting the assumption that the
%   construction works for an arbitrary $m$. Hence $b < a$.

%   Second, we can show that if $b = a \land \bigwedge_{i = 1}^k \phi_i$ for some
%   positive integer $k$, then there must also be $2^k - 1$ other atoms in
%   $\mathbf{B'}$ that correspond to every possible way to negate a subset of
%   $\phi_i$'s, i.e., ranging from

%   % 2. If we add phi, then we must also add -phi.
%   % 3. Extension to multiple literals: we must have all (2^n) combinations of
%   % added literals).
%   % 4. Profit
% \end{proof}

Let us note how our lower bound on the number of added literals compares to two
methods of translating a discrete probability distribution into a WMC problem
over a propositional knowledge base proposed by Darwiche
\cite{DBLP:conf/kr/Darwiche02} and Sang et al. \cite{sang2005solving}. Suppose
we have a discrete probability distribution with  $n$ variables, and the $i$-th
variable has $v_i$ values, for each $i \in [n]$. Interpreted as a logical
system, it has $\prod_{i=1}^n v_i$ models. My expansion would then use
\[
  \sum_{i=1}^n v_i + 2\prod_{i=1}^n v_i
\]
variables, i.e., a variable for each possible variable-value assignment, and two
additional variables for each model. Without making any independence
assumptions, the encoding by Darwiche \cite{DBLP:conf/kr/Darwiche02} would use
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n \prod_{j=1}^i v_j
\]
variables, while for the encoding by Sang et al. \cite{sang2005solving},
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n (v_i - 1) \prod_{j=1}^{i-1} v_j
\]
variables would suffice.
